### 2012-02-13 doc/theses/mikula/pre_creation_of_algorithm.tex
Master thesis: In progress.

@@

The result of lexical and syntax analyses can be some kind of so-called syntax tree \cite{compilers}. It is a structure representing a word according to a formal grammar of some language. In our case, the language is XQuery, its grammar is defined in REF \todo{ref} and every query is a word of the XQuery language. Leaves of the tree represent terminals of the grammar while inner nodes represent non-terminals. \todo{priklad nejakeho stromu?} From the point of view of this work, the syntax tree can be perceived as a preprocessed form of a query keeping its complete meaning and making its further processing more convenient. For instance, the tree can simplify a search for FLWOR statements. It is transitioned and nodes representing FLWORs are found. Then each subtree determined by one of the found nodes represents a FLWOR statement and it can be analyzed further.

@@

The following text is an example of inference of a more complex query processing. Consider the following part of a query.
\begin{verbatim}
declare function local:getB($id as xs:string) as element() {
  //A[@id = $id]/B
}
... local:getBs("id") > 10 ...
\end{verbatim}
The query consists of a function declaration and an arithmetic comparison. The comparison compares a result of the function call and literal value \texttt{10}. For the type of \texttt{10} is \texttt{xs:integer} the type of the function call has to be convertible to \texttt{xs:double}. Thus, it has to be a numeric type. The function returns a path expression typed as \texttt{element()}. That means the function returns one element \todo{pravda?}. In the path expression, the argument \texttt{\$id} can be substituted by the real value specified in the call. Thus, the return expression is \texttt{//A[@id = "id"]/B}. Therefore, we can infer that element \texttt{B} in element \texttt{A} with attribute \texttt{id} equaling string value \texttt{"id"} is of some numeric type. And, for the function is parametrized there is a notion that this statement may be correct also for other elements \texttt{A} and \texttt{B}.

@@

\todo[inline]{Tak náročnost implementace rozhodně není kritérium. Ta se udělá jednou. Navíc ve vědě (tj. i DP) nemá opodstatnění. Pro DP navíc platí, že nemusíte implementovat vše, co navrhnete. Implementace je pomocná věc, která slouží k tomu, abychom experimentálně ukázali vlastnosti a chování navrženého algoritmu.}

\section{Inference of XML Structure}
The question of inference of XML structure from queries is partially discussed in the previous chapter. We are able to infer XML structure from queries without their evaluation, but in a limited way. This inference is based on an analysis of path expressions. Its limitation involves the following issues. When we infer some subelements of a certain element we often cannot be sure about their number of occurrences. Also, we cannot be sure if every occurrence of the certain element contains these subelements and we even do not know if at least one occurrence of the element contains them. Thus, the inferred statement is more likely an indication than a fact about the structure.

@@

\section{Extension of an Existing Approach}
The existing approaches of XML schema inference deal mainly with the inference of XML structure. As stated in the previous section, this work does not deal with the inference of XML structure, and hence, the extension of an existing approach will be a kind of an independent addition instead of modification and refinement of its core algorithm.

@@

>>>>

\todo[inline]{Určitě to chce kombinaci dotazů a dat. Na začátku textu říkáme, že samotné dotazy nestačí, resp. neříkají všechno. Ale klidně můžete v textu popsat oboje a pak implementovat jen část z obojího a porovnat výsledky. Nebo vzít nějakou metodu, která je v jInfer (DP p. Klempy), rozšířit ji o zpracování dotazů a na nějakých pěkných datech ukázat, že je to rychlejší/dává lepší (přesnější) výsledky.}

@@

The result of lexical and syntax analyses can be a kind of so-called syntax tree \cite{compilers}. It is a structure representing a word according to a formal grammar of a language. In our case, the language is XQuery, its grammar is defined in REF \todo{ref} and every query is a word of the XQuery language. Leaves of the tree represent terminals of the grammar while inner nodes represent non-terminals. \todo{priklad nejakeho stromu?} From the point of view of this work, the syntax tree can be perceived as a preprocessed form of a query keeping its complete meaning and making its further processing more convenient. For instance, the tree can simplify a search for FLWOR statements. It is transitioned and nodes representing FLWORs are found. Then each subtree determined by one of the found nodes represents a FLWOR statement and it can be analyzed further.

@@

The following text is an example of inference of a more complex query processing. Consider the following part of a query.
\begin{verbatim}
declare function local:getB($id as xs:string) as element() {
  //A[@id = $id]/B
}
... local:getBs("id") > 10 ...
\end{verbatim}
The query consists of a function declaration and an arithmetic comparison. The comparison compares a result of the function call and literal value \texttt{10}. For the type of \texttt{10} is \texttt{xs:integer} the type of the function call has to be convertible to \texttt{xs:double}. Thus, it has to be a numeric type. The function returns a path expression typed as \texttt{element()}. That means the function returns one element \todo{pravda?}. In the path expression, the argument \texttt{\$id} can be substituted by the real value specified in the call. Thus, the return expression is \texttt{//A[@id = "id"]/B}. Therefore, we can infer that element \texttt{B} in element \texttt{A} with attribute \texttt{id} equaling string value \texttt{"id"} is of some numeric type. And, for the function is parametrized \todo{divná věta} there is a notion that this statement may be correct also for other elements \texttt{A} and \texttt{B}.

@@

\section{Inference of XML Structure}
The question of inference of XML structure from queries is partially discussed in the previous chapter. We are able to infer XML structure from queries without their evaluation, but in a limited way. This inference is based on an analysis of path expressions. Its limitation involves the following issues. When we infer some subelements of a certain element we often cannot be sure about their number of occurrences. Also, we cannot be sure if every occurrence of a certain element contains the subelements and we even do not know if at least one occurrence of the element contains them. Thus, the inferred statement is more likely an indication than a fact about the structure.

\todo[inline]{S proč potřebuju vyhonotit dotazy? Dostanu nějaký výsledek, ale ten může obsahovat třeba úplně jiná data ne jsou v dokumentech - použiju konstruktory, něco sečtu a původní idokumenty můžou být úplně jiné.}

@@

\todo[inline]{Ta věta je nejasná. Spíš pro to odvození struktury použijeme některý standardní přístup analyzující data a dotazy použijeme pro další upřesnění, ne? Nebo strukturu nebudeme řešit vůbec? To je asi divné.}

\section{Extension of an Existing Approach}
The existing approaches of XML schema inference deal mainly with inference of XML structure. Hence, the extension of an existing approach will be a kind of an independent addition instead of modification and refinement of its core algorithm.

@@

\todo[inline]{Také bychom mohli vzít přístup odvozující strukturu a informace z dotazů využít v  jeho průběhu. Obecně tyto metody vytvoří prefixový strom z iniciální gramatiky a pak se ho snaží zobecňovat sléváním stavů a zobecňováním automatu. Např. můžeme mít pravidlo, že pokud se element A vyskytuje alespoň 5x, pravděpodobně se může vyskytovat libovolněkrát. My ale v datech najdeme element jenom 4x. Pak nevím jeslti je pravidlo rozumné. Dotazy nám to mohou potvrdit/vyvrátit. Čili můžeme zpřesnit proces odvozování.
Nebo ho můžeme zoptimalizovat. Např. Vošta ve své diplomce poměrně složitě analyzuje automat aby našek konstrukt xs:all. My bychom to pomocí analýzy dotazů mohli umět jednodušeji = efektivněji.
Nebo jsme schopni pomocí dotazů přesněji určit kontext elementu (opět řeší Vošta v diplomce). Např. mám element jméno, který se vyskytuje v různých kontextech (absolutních cestách). Já ale nevím, jeslti ten element má stený význam a tudíž by pro něj měl být odvozen stejný model obsahu. Např. jméno autora může být složeno z podelementu křestní a příjmení, zatímco jméno knihy bude prostý string. Vošta používá podobnostní funkci, ale tam je potřeba nastavit threshold. Pokud ale budu mít dotaz, který mi tyto elementy dává do jedné proměnné (např. jméno zákazníka i jméno autora), zřejmě jsou sémanticky podobné a mohou mít stejný model obsahu. Stejný dotaz pro knihy i autory by byl asi divný. 
Tohle by asi nebylo špatné minimálně odiskutovat. Mrkněte se např. na diplomku Vošty (popř. Klempy). Jsou tam popsané typické postupy pro odvozování a základní pravidla slévání. U nich by také dotazy mohly pomoci.}

### 2012-02-13 doc/theses/mikula/proposed_algorithm.tex
Master thesis: In progress.

The syntactic analysis needs to be slightly modified to suit our case. It writes its result into a file in an XML representation. Instead, we need to keep the result in memory and pass it to consecutive steps of our algorithm. Though this requirement concerns modifications of implementation, the core of the processing remains untouched. Therefore, we also do not describe these modifications.

@@

\begin{define}[Syntax tree]
\label{DEF_syntax_tree}
Syntax tree of XQuery query $Q$ is tuple $T = (V, E, c, \mathcal{P}, o)$ where $V \subset \mathbb{N}$ is a set of nodes, each node representing a particular XQuery construct in query $Q$, $E$ is a set of pairs $(v, w)$ where $v,w \in V$ and for every $a, b \in V, a \neq b: (a, b) \in E$ if and only if a construct represented by $b$ is a direct component of a construct represented by $a$ ($b$ is a child of $a$) in query $Q$, $c: V \rightarrow C$ is function assigning each node with its class from set $C$ of all XQuery language constructs listed in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}, $\mathcal{P}$ is a set of functions specifying additional properties of the nodes and distinguishing the nodes of the same classes, and $o : V \rightarrow \mathcal{O}$ is a function specifying an order of children of the nodes where $\mathcal{O} = \{o_v : E_v \rightarrow \mathbb{N} | v \in V, E_v = \{(v, w) | w \in V, (v,w) \in E\}\}$ is set of functions specifying the children order for each node. For every $v \in V, o(v) = o_v$ so that $o_v(v,w)$ is a sequential number of a construct represented by $w$ amongst constructs represented by children of $v$ in query $Q$. 
\end{define}

@@

Set $\mathcal{P}$ contains other similar functions but due to their large count, we do not define them formally.
\todo[inline]{Moze takto ostat, alebo definovat aspon tie co pouzivame? Alebo ich aspon vypisat?}

@@

\todo[inline]{Je nutne presnejsie popisat kedy a preco tieto skratky maju zmysel? V pseudo-kodoch budu samozrejme pouzivane len v pripadoch, ked su dobre definovane a maju zmysel.}

\subsection{Closer Look at the Nodes of the Syntax Tree}
\todo[inline]{Dat figures do appendixu?}
The node classes of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where an object can be of several types. This hierarchy is shown in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}. The classes with emphasised names represent a common class of a group of subclasses, and these classes cannot be directly used in the syntax tree (abstract classes). The remaining nodes represent particular constructs of the XQuery language and nodes of these classes can be used in the syntax tree.

@@

\todo[inline]{Asi bude potrebne nakreslit pre kazdy node, co presne obsahuje. Alebo staci pre tie, ktore budeme my pouzivat?}

@@

Some pseudo code algorithms in the following sections need to determine a class of a node. The node's direct class can be determined by function $c$ from the definition of the syntax tree \ref{DEF_syntax_tree}, and hence, we also can determine its indirect classes. For the purpose of pseudo code, we define the following function. For every $v \in V$ and every $class \in C$, function $is(v, class)$ returns boolean value $true$ if $c(v) = class$ or $c(v)$ is a (direct or indirect) subclass of $class$, according to the described principle. Otherwise, it returns $false$.

@@

\subsection{Characteristics of the Syntax Tree}
An important characteristic of the syntax tree is related to a definition of local variables and their scope in the XQuery language. The representation of a definition of a local variable in the syntax tree is a node of $VariableBinding$ class. Nodes of that class contain have only two children; a node representing the type of the variable and a node representing the binding expression (expression defining the value of the variable, and thus, it cannot use the variable). Hence, the entire subtree doesn't contain any expressions that use the variable. Therefore the scope of the new variable is not the subtree of the $VariableBinding$ class node. It depends on the type of XQuery construct that the variable binding is a (indirect) component of.

For example, a node of $FLWORExpr$ class contains four subnodes of $TupleStream$, $WhereClause$, $OrderByClause$ and $ReturnClause$ classes. The $TupleStream$ class node contains a list of node of $variableBinding$ class which define variables valid in all other three subnodes of the FLWOR node.

@@

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Types $ElementType$, $AttributeType$, $NodeType$, $TextNodeType$, $CommentType$, $ProcessingInstructionType$, $DocumentType$ representing an element, attribute, node, text node, comment, processing instruction, document node.
\item Type representing a node selected by a certain path expression. The path expression is included in this type. Let this type be identified as $PathType$.
\end{itemize}

@@

$PathType$ has always cardinality of zero-or-more, because the result of a path expression is a general sequence and since we do not evaluate the queries we cannot determine its actual cardinality.

@@

\todo[inline]{Este by sa dala urobit analyza volania funkcii, do ktorych su predavane nejake elementy/atributy o ktorych vieme ich path expression. Vysledkom tejto funkcie by totiz zas mohli byt elementy/atributy o ktorych vieme (pozmenenu) path expression a nie len nezname elementy/atributy.
A: Přesně tak. A jak říkám - není třeba vše co navrhnete implementovat.}

\subsection{Determination of Function Return Types}
Determination of return types of functions is needed because function calls can appear in expressions. A return type of a function can be determined at the moment the analysis of expressions encounters a call of the function, however, it involves multiple transitions of the syntax tree in a search for a definition of a particular function.

@@

A simple algorithm is presented in Algorithm \ref{ALG_processing_of_functions}. It uses \texttt{getFunctionDeclarationNodes()} function, defined in Algorithm \ref{ALG_retrieval_of_function_declaration_nodes}, which returns a list of all function declaration nodes in the syntax three. Actually, it does not have to search the whole syntax tree as the function declaration nodes can be present only in the query prolog section.

Without loss of generality, we will focus on locally defined functions. We will not determine types of functions defined in other modules since the principle is similar but it is needed to look up the definitions in syntax trees of other queries. Return types of built-in functions are fixed and, thus, there is no need to analyse them.

>>>>

The syntactic analysis needs to be slightly modified to suit our case. It writes its result into a file in an XML representation. Instead, we need to keep the result in the main memory and pass it to consecutive steps of our algorithm. Though this requirement concerns modifications of implementation, the core of the processing remains untouched. Therefore, we also do not describe these modifications.

@@

\begin{define}[Syntax tree]
\label{DEF_syntax_tree}
Syntax tree of XQuery query $Q$ is tuple $T = (V, E, c, \mathcal{P}, o)$ where
\begin{itemize}
\item $V \subset \mathbb{N}$ is a set of nodes, each node representing a particular XQuery construct in query $Q$,
\item $E$ is a set of pairs $(v, w)$ where $v,w \in V$ and for every $a, b \in V, a \neq b: (a, b) \in E$ if and only if a construct represented by $b$ is a direct component of a construct represented by $a$ ($b$ is a child of $a$) in query $Q$,
\item $c: V \rightarrow C$ is function assigning each node with its class from set $C$ of all XQuery language constructs listed in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}, \todo{Tohle je hrozné peklo. Potřebujeme to takto? Nedalo by se to nacpat do 1-2 tabulek/obrázků? Je nutné aby každá položky byla na samostatném řádku. Trochu to vypadá jako umělé natahování diplomky.}
\item $\mathcal{P}$ is a set of functions specifying additional properties of the nodes and distinguishing the nodes of the same classes,
\item and $o : V \rightarrow \mathcal{O}$ is a function specifying an order of children of the nodes, where $\mathcal{O} = \{o_v : E_v \rightarrow \mathbb{N} | v \in V, E_v = \{(v, w) | w \in V, (v,w) \in E\}\}$ is set of functions specifying the children order for each node. For every $v \in V, o(v) = o_v$ so that $o_v(v,w)$ is a sequential number of a construct represented by $w$ amongst constructs represented by children of $v$ in query $Q$.
\end{itemize}
\end{define}

@@

Set $\mathcal{P}$ contains other similar functions but due to their large number, we do not define them formally.
\todo[inline]{Q: Moze takto ostat, alebo definovat aspon tie co pouzivame? Alebo ich aspon vypisat? A: aspoň příklad, lépe vypsat ty, co použijeme}

\todo[inline]{Hodil by se příklad stromu - jeslti nebude dál.}

@@

\todo[inline]{Q: Je nutne presnejsie popisat kedy a preco tieto skratky maju zmysel? V pseudo-kodoch budu samozrejme pouzivane len v pripadoch, ked su dobre definovane a maju zmysel. A: Ne, ale hodilo by se slovně popsat co znamenají.}

\subsection{Closer Look at the Nodes of the Syntax Tree}
\todo[inline]{Dat figures do appendixu?}
The node classes of the syntax are organized in an is-a hierarchical structure, commonly used in the object oriented programming languages, where an object can be of several types. This hierarchy is shown in Figures \ref{FIG_syntax_tree_node_types_part_1}, \ref{FIG_syntax_tree_node_types_part_2}, \ref{FIG_syntax_tree_node_types_part_3}, \ref{FIG_syntax_tree_node_types_part_4}, \ref{FIG_syntax_tree_node_types_part_5}, \ref{FIG_syntax_tree_node_types_part_6}, \ref{FIG_syntax_tree_node_types_part_7}, \ref{FIG_syntax_tree_node_types_part_8}, \ref{FIG_syntax_tree_node_types_part_9}. The classes with names in bold represent a common class of a group of subclasses, and these classes cannot be directly used in the syntax tree (abstract classes). The remaining nodes represent particular constructs of the XQuery language and nodes of these classes can be used in the syntax tree.

@@

\todo[inline]{Q: Asi bude potrebne nakreslit pre kazdy node, co presne obsahuje. Alebo staci pre tie, ktore budeme my pouzivat? A: Určitě všude stačí popsat jen to, co použijeme, s poznámkou, že zbytek je např. v diplomce, ze které vycházíte.}

@@

Some pseudo code algorithms in the following sections need to determine a class of a node. The node's direct class can be determined by function $c$ from the definition of the syntax tree in Definition \ref{DEF_syntax_tree}, and hence, we also can determine its indirect classes. For the purpose of pseudo code, we define the following function.

\begin{define}[Function is(v, class)]For every $v \in V$ and every $class \in C$, function $is(v, class)$ returns boolean value $true$ if $c(v) = class$ or $c(v)$ is a (direct or indirect) subclass of $class$, according to the described principle. Otherwise, it returns $false$.
\end{define}

@@

\subsection{Characteristics of the Syntax Tree}
An important characteristic of the syntax tree is related to a definition of local variables and their scope in the XQuery language. The representation of a definition of a local variable in the syntax tree is a node of $VariableBinding$ class. Nodes of that class have only two children; a node representing the type of the variable and a node representing the binding expression (expression defining the value of the variable, and thus, it cannot use the variable). Hence, the entire subtree does not contain any expressions that use the variable. Therefore the scope of the new variable is not the subtree of the $VariableBinding$ class node. It depends on the type of XQuery construct that the variable binding is a (indirect) component of.

For example, a node of $FLWORExpr$ class contains four subnodes of\linebreak $TupleStream$, $WhereClause$, $OrderByClause$ and $ReturnClause$ classes. The $TupleStream$ class node contains a list of node of $variableBinding$ class which define variables valid in all other three subnodes of the FLWOR node.

@@

\begin{itemize}
\item XML Schema built-in atomic types \todo{link do appendixu}.
\item Types $ElementType$, $AttributeType$, $NodeType$, $TextNodeType$,\linebreak $CommentType$, $ProcessingInstructionType$, $DocumentType$ representing an element, attribute, node, text node, comment, processing instruction, document node.
\item Type representing a node selected by a certain path expression. The path expression is included in this type. Let this type be identified as $PathType$.
\end{itemize}

@@

$PathType$ has always cardinality of zero-or-more, because the result of a path expression is a general sequence and since we do not evaluate the queries, we cannot determine its actual cardinality.

@@

\todo[inline]{Este by sa dala urobit analyza volania funkcii, do ktorych su predavane nejake elementy/atributy o ktorych vieme ich path expression. Vysledkom tejto funkcie by totiz zas mohli byt elementy/atributy o ktorych vieme (pozmenenu) path expression a nie len nezname elementy/atributy.
A: Přesně tak. A jak říkám - není třeba vše co navrhnete implementovat.
A2: Stačí to popsat a v kapitole Implementace říct, které ásti algoritmu implementujeme. Stačí takové, které už ukáží zajímavé chování a dokážou přínosy návrhu. Přesně k tomu implmentace je.}

\subsection{Determination of Function Return Types}
Determination of return types of functions is needed because function calls can appear in expressions. A return type of a function can be determined at the moment the analysis of expressions encounters a call of the function; however, it involves multiple transitions of the syntax tree in a search for a definition of a particular function.

@@

A simple algorithm is presented in Algorithm \ref{ALG_processing_of_functions}. It uses \texttt{getFunction\-DeclarationNodes()} function, defined in Algorithm \ref{ALG_retrieval_of_function_declaration_nodes}, which returns a list of all function declaration nodes in the syntax three. Actually, it does not have to search the whole syntax tree as the function declaration nodes can be present only in the query prolog section.

Without loss of generality, we will focus on locally defined functions. We will not determine types of functions defined in other modules since the principle is similar but it is needed to look up the definitions in syntax trees of other queries. Return types of built-in functions are fixed and, thus, there is no need to analyze them.
